Towards the Publication of ICI Siracusa: General Data and Previews.

### 2016-01-13 source/32/EAGLE_2016_submission_32.tex
paper 32 convertito
bibliografia aggiunta in biblio, riferimenti nel testo aggiornati

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Towards the Publication of ICI Siracusa: General Data and Previews.}
\author[unica]{Mariarita Sgarlata\corref{first}}
\address[unica]{Department of Humanities - University of Catania} 
\cortext[first]{Corresponding author. Email: m.sgarlata@unict.it}
\begin{document}
\maketitle
\begin{abstract}
This paper refers to the general track “Epigraphic edition on paper and on line”. Some researchers who has been working on the editing of the Inscriptiones Christianae Italiae, published from the University of Bari, contributed also to the EDR project, that collect on line the epigraphic documents of Roman Christian Period. I propose a preview of the work currently in progress, with a specific reference to the inscriptions that provide us the chronological and topographic data to study the cemetery as well as formularies linked to the society structure and to the identity-making characteristics.
\end{abstract}
\keywords{Sicily, Syracuse, Epigraphy, Topography, Society, Onomastic and Identity}

@@

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{MISSING IMAGE!}
\caption{The main gallery (so called decumanus) of Catacombs of San Giovanni}
\label{fig:1}
\end{figure}

We are far from a complete and thorough research about palaeochristian epigraphs in Sicily. This catalogue represents the first step of a systematic study focused on the inscriptions discovered in the christian cemeteries of Siracusa.

@@

The epigraphic research related to San Giovanni catacomb can count on certain informations about the discovery place thanks to Paolo Orsi archeological campaigns. However, looking to previous publications, sometimes it’s necessary a review of the presentations made by Mommsen and Kaibel. In this perspective the main sources of the CIL and IG authors were verified so we had a deeper vision of the relationship between Mommsen and Kaibel and the main sources used, sometimes reevaluating the contribution of this sources. The renovated interest about funerary epigraphy generated new studies about new criteria of dating. The inscriptions dated by the indications of the ‘’consular couple’’ represent an exceptions in the issues about chronological datings (FERRUA 1947, 227; 1989, nn. 191-208): two were discovered during the excavation conducted by Paolo Orsi (1985) and to them is associated the latin epigraph ‘’Sporus’’ (356) .
The bigger one is entitled to San Giovanni and gave back almost 800 inscriptions, now located for the most part in the Archeological and Regional Museum ‘’Paolo Orsi’’ of Siracusa.; alongside will be placed the 40 headstones recovered in the adjacent hypogea of Villa Landolina, that report compatible datings and formularies.

1.2 HISTORICAL ASPECTS OF THE CATACOMB OF S. GIOVANNI AT SYRACUSE
The systematic studies of collective cemeteries in Syracuse started many years ago and, beyond the archaeological surveys, the late antique funerary settlements of Syracuse proposed multiple research cues and paths, as the historical- religious, economic and social nature.
Suburban cemeteries, fanned out from the area of Fusco, in the quarter of Neapolis, to the Santa Lucia area, in the southern part of Acradina, this indicates unequivocally what the perimeter of the city must have already been in the early and mid Roman Empire. The History of the area, which was going to hold the catacombs (San Giovanni, Vigna Cassia and Santa Lucia), spanned the centuries between the classical Greek and late antique ages, gradually giving evidence of quarries (Latomie), water supply systems to the city, characterized by cisterns and aqueducts (COLLIN BOUFFIER 1987, 682), handcraft workshops from the beginning of the 4th/3rd century BCE and burials datable to the early and mid Roman Empire. The analysis of the funerary system certifies one hand the dependence on the Roman model, and other the debt in respect of local traditions.
Several interest will be given to structural aspect of the catacomb of S. Giovanni, practice of funeral rituals, ethnic and cultural fruition’s characters, transformation in the use, transformation in the way of using spaces for graves, to complete a general point of view about the phenomena of continuity and innovation as to previous sepulchral arrangements and, in the analysed periods, the facies belonging to the different settling, variegated in the committees’ideological and religious themes, in choosing monumental types (like the rotundae) and decorations, in self-representative aspects, in burial uses. In this perspective we will give particular importance to the study of funerary epigraphy aimed at the writing of ICI Siracusa and the overall interpretation of the monument.
Between 1893 and 1909 the archaeologist Paolo Orsi carried out a series of campaigns in the catacomb of San Giovanni. Detailed reports of those campaigns are recorded at various times in Notizie degli Scavi (see the BIBLIOGRAPHY), which constitute an indefeasible starting point. The first incisive studies on Christian subterranean Sicily pertain to Joseph Führer, who dedicates many pages to the catacomb of San Giovanni, whose study is also epigraphic, is based on the previous literature, especially on Paolo Orsi’s discoveries (FÜHRER 1897, 13-39; FÜHRER- SCHULTZE 1907, 22-26; see also ORSI 1893, 276, n. 2 and SCHULTZE 1882, 60-61, fig. 5). In the study of Syracuse cemeteries Orsi (ORSI 1900a, 189) was the first to see the mixed nature of the burials and their materials (mostly gravestones), openly attesting a kind of pagan-Christian and orthodox-heterodox symbioses (see also AGNELLO 1957a, 242-243). Both Antonio Ferrua and Santi Luigi Agnello within the space of a few years tried to solve the problems relating to the sarcophagus of Adelfia and the more general problem, strictly correlated with the ones just mentioned, of the cemetery genesis and development (FERRUA 1952 and AGNELLO 1956). It is necessary to record that numerous hypotheses on the genesis and development of the various sectors of the cemetery are ascribable to the account given by inscriptions. Syracuse, thanks to its prolific underground cemeteries, has the larger Christian epigraphic heritage after Rome, kept in the storage of the Soprintendenza di Siracusa.
1.3 TOPOGRAPHIC AND EPIGRAFIC DATA TO STUDY THE CEMETERY
Walking across the main gallery, one can retrace the stages of Paolo Orsi’s interventions, being at the entrance of the Catacomb second northern gallery, before the so-called “tomb of the Saint”. This arcosolium is considered a privileged burial on the basis of the following reasons: most of all because of its position, but also because it is a mensatype burial closed by a sole slab (ORSI 1893, 292-294). The signs of an ancient rite are easily traceable on the slab, a rite that preceded the coming of Christianity and persisted for centuries up to the present day: the rite of refrigerium, which literally means refreshment, cooling (GIUNTELLA 1985). In the Christian ceremony the purpose of a funeral banquet is to benefit the soul of the departed on the anniversary of death, a painful event celebrated as dies natalis of the soul to eternal life.
Who was buried in this sepulchre? The question is destined to remain unanswered and only an inscription found nearby could be a clue in this sense. The text says that the owners purchased the sepulchre close to the one of the bishop Cheperion’s (ORSI 1895, 507-508; RIZZONE 2011, PP. 55-58), of whom the scanty written sources never make mention. In any case this constructed sarcophagus represent one of the most eminent burials of the catacomb. In the same gallery one cannot ignore the finding of an inscription in many fragments, which is a singular phenomenon of religious contamination.
This inscription, with a Christological monogram at the top, records Nassiana “Christian, who competed with Penelope in moral virtue”. This inscription can only be compared with three examples in the Roman sepulchral carmina (BÜCHELER 1897). The circular support of this inscription found close to the “tomb of the Saint” has been regarded as a
mensa(table) for refrigerium rite (GIUNTELLA 1985, 47), whose circular form would derive in any case from an evident reuse of a marble disk of classical craftsmanship, with a laurel wreath and berries sculpted on one side (ORSI 1895, n. 234). A wall inscription painted on the extrados of an arcosolium in the third northern gallery of the catacomb could be connected with Nassiana’s text. Two lines of the inscription say that “Sossa outdid (other women) in conjugal love; as for handiworks, without being taught by anyone, Athena herself had taught her how to do marvellous things” (FERRUA 1940, n. 3). Both inscriptions unequivocally make use of female figures from the classical world, interpreted as a model also for Christian women either for a deep-rooted usage or making up for the absence of assimilable figures in the Church. This allows one to notice in a 4th century large community cemetery, such as San Giovanni, some phenomena of religious contamination in a period in which Christian epigraphic praxis consolidated itself by then (SGARLATA 1999, 484).
The rotunda derives its name from the deceased Antiochia recorded in the sarcophagus, set inside the ring of tombs, made of blocks and bearing an engraved and rubricated inscription. Whom did this private space belong to? To answer the question, one can rely on a suggestion provided once more by epigraphic testimonies. The gravestones found – on which they put as a rule the deceased’s name, lifespan, date of death and deposition – according to the first excavators’ reports (CARINI 1873, 23-45), attest that the rotunda of Antiochia could have housed women only and this would give credence to the idea that the mausoleum had been used for a female monastic community. The hypothesis needs to be verified, but it is very seductive.
It is evident that the catacomb of San Giovanni was originally a community cemetery, planned for only one type of burial: the arcosolium with multiple depositions, which does not require great care for decoration, only transennae. In the topographical and architectural development of the catacomb it appears clear that creating the rotundas breaks up the common burials series, destined to a socially homogeneous Christian community. These changes to the original plan – of the creation of subterranean mausolea both to the north and south of the cemetery – spring from the necessity to create appropriate spaces for the members of the Chruch and above all of the Empire, bringing into question the initial egalitarian choice of the arcosolium burials (GRIESHEIMER 1989, 767). In the terminal part of the pious Giovanni’s gallery a monumental sarcophagus, once more hewn out of the rock, can be perceived. Close to the closing wall of the same gallery an inscription was found: it bears the notification of both the consuls of the year 349 (ORSI 1909, 756, 8; AGNELLO 1953, 89). As a rule this date marks the end of the dig works in this sector of the catacomb (FERRUA 1952, 75- 76). A arcosolium extrados, in the eastern region, shows traces of a palinsesto, an overlap of paintings and epigraphs than obstruct the identification of the first person buried in this sepulchre, even if recently Vittorio Rizzone proposed the interpretation Philadelpheia1. In one of the fossa tombs at the terminal stretch of the main gallery the inscription of Euterpe (IG XIV, 112) has been found, reused to cover the bottom of the tomb, which traditionally marks the end of the dig works of the catacomb (FERRUA 1952, 75-76; AGNELLO 1958, 79; GRIESHEIMER 1989, 781). Beyond the deceased’s biometric data, recorded as “companion of the Muses”, the epigraph in the last three lines mentions the consuls’ iteration, which allow us to date back to the consulate of the Emperor Constantius, consul for the tenth time, and Julian- Caesar, consul for the third time. So Euterpe died on November 27th 360 A.D. at the age of 22 years and 3 months (GUARDUCCI 1978, 524-526).
In the southern region,, more than in other ones, that you can note the transformations that have profoundly undermined its community spirit, which had originally inspired the creation of the catacomb. In the first rotunda of the southern region, is a private space, which was given the name of Marina due to an inscription scratched upon the extrados of the arcosolium,on the right of the entrance to the short gallery of the bishop Siracosio. The arcosolium seems to be enframed by a painted prothyrum, as the still visible column and capital attest, confirming the generalized use of architectural elements in this catacomb, already seen in the rotunda of Antiochia. According to the interpretation of the text, Marina could have been the wife of the patriciuset magister militum Sabinianus, sent by the Emperor Honorius to Spain at the time of barbarian invasions presumably between 409 and 423 (FERRUA 1989, 21-22, 40).
This date would agree with the numerous testimonies given by the inscriptions that have been dated thanks to the notification of the two consuls in office in the year of their death.These testimonies are ascribable to the years of the Emperors Arcadius and Honorius in the first quarter of the 5th century and seem to suggest a link between a few burials of the southern region and the aristocrats’ diaspora from Rome after Alaricus’ advance in 410, who took refuge in Sicily and Africa as they did in other provinces of the Empire (SIRAGO 1989, 175; VERA 1988, 157-158). It is necessary to have a look at the slab of the presumed arcosolium of the bishop Siracosio, recorded in one inscription found in an adjacent fossa tomb, which says that the deceased intentionally purchased the sepulchre close to the one of the bishop just mentioned (IG XIV, 123: Ἐνθάδε κῖτε Πολυχ / ρόνιος καὶ Σεραπία. / Ἠγόρασεν τῷτότ / ε καιρῷ / Πολυχρονίου /
 1 RIZZONE 2012, pp. 45-48.
αἱ Σεραπία ἐπὶ τῷκυρί / ῳ μου ἐπισκόπῳ Συρα / κοσίῳ). It is just a hypothesis seeing in the arcosolium with the engraved slab, still in situ, a noble burial for a member of the ecclesiastical hierarchy, for whom, in the absence of other data, according to a letter of pope Gelasius I, an episcopate between 492 and 496 has been proposed (NARCISO 1952, 223, CARLETTI 2008). One can clearly distinguish a Christogram with the apocalyptic letters alpha and omega and two ships in the shape of fish, regarded as making reference to the sacrament of the Eucharist, this is also suggested by the diskettes next to the fishes’ mouths, assimilable to loaves of bread (SGARLATA 2013).
After reading the text of the inscription upon the sarcophagus lid (CIL X, 7123: Ic Adelfia c(larissima) f(emina) / posita conpar / Balericomitis), one can become aware that it refers only to the woman, not to both husband and wife: here lies Adelfia, clarissima femina, wife of the count Valerius. Who were therefore Valerius and Adelfia? The unsolved enigmas of the catacomb of San Giovanni deal with their names and most of all their identification. In reality it is possible to distinguish more than two phases of intervention, which have preceded the creation of the hole for the sarcophagus. At the beginning the internal space of the large niche was scenographically arranged in a terrace pattern with a forceful ascensional effect, not exempt from comparisons (NESTORI 1973, 14, plate VII).The intact Latin inscription of Sporus, dated to 356, which sealed a forma in the main gallery among the ones that join the rotunda of Marina to the one of Adelfia (CAVALLARI 1872, 24; AGNELLO 1953, 90), would attest that the exploitation of the soil in the area gravitating to the rotunda of Adelfia was already underway after the first half of the 4th century, so confirming the evident anteriority of the six graves cut in the floor to the phase of monumentalization. Both the internal and the external organizations of the large niche (nicchione) would correspond to the first and second phases of intervention respectively, according to the current reconstruction. The problems related to the reconstruction of the third phase seem to be more complicated: the monumentalization phase started with the interment of the sarcophagus and concluded with the acquisition of an aspect comparable to the privileged burials of Roman crypts.
The monument seems to have a less wild internal dynamics of development; the new data (archaeological, historical, epigraphical) (SGARLATA 1996, 101-108) allow the scholars to widen the syncopated weave of temporal sequences, to which the analysis of the monument has been pinned by the constant reference to Valerius Proculus’ chronology. A new chronology and a different identification of the comes Valerius can be proposed, considering: 1) the evident reuse of the sarcophagus; 2) the topographical development of the catacomb in the area where the sarcophagus has been discovered; 3) the type of monumental intervention, which followed the Roman counterparts, datable to the second half of the 4th century (FIOCCHI NICOLAI 1997, 132-134). If Valerius were given a different identity, one could postpone him from the age of Constantine to that of Augustine (SGARLATA 1998, 15-51), on the basis of several accounts about the friendship between the saint from Hippo and a comes Valerius, whose physiognomy, is somewhat vague. This friendship, confirmed by the epistles (AUG., Ep. 200, 206, 207; Retr. II, 79 and 88) and the dedication, in 419, of the treatise De nuptiis et concupiscentia, is fed on the fight against Pelagianism, which in eastern Sicily had found a fertile soil. The presence of Pelagians in Sicily is attested for certain, according to Hilarius’ account under the pontificate of Innocent (PIETRI 2000, 429-452) and Honoreficentia’s letter, where a clarissima based at Syracuse is mentioned. On the island the spread of Pelagian movement looks like a direct consequence of the 410 sacking of Rome and the diaspora of the Roman nobility, of which Pelagius and Celestius were spiritual leaders; the short period they both spent in Sicily was not painless for Christian orthodoxy (PIETRI 2001, 436-437) and particularly in Syracuse, as the cemeteries in the area overhanging the Greek theater, intended to serve the communities of the so-called heretics throughout the 5th century (AGNELLO 1990).All the data gathered seems to lead to a different identification of Valerius – who, even if he was not Augustine’s correspondent, would be sought in the list of Valerii reported in the sources of the first quarter of the 5th century (PLRE II, 1143-44) – and a later chronology of the large niche monumental transformation compared to the one traditionally accepted.

@@

To answer the questions will take time, which will assure a greater credibility to what so far seems just a series of suggestions, most of all fed by the re-reading of epigraphic testimonies: the episodic character of the references to burials of bishops, presbyters, deacons and noting that the most significant percentage of evidences pertains to the members of the Church, buried in Syracuse away from their own countries and recorded in wall inscriptions written in Latin – Auxentius Hispanus episcopus and Superianus clerecus de Aquileia (FERRUA 1940, 1 and 6) – as a demonstration that the official language is used by foreigners, who were high clients (the remainder of inscriptions is in Greek), these and many other clues lead to thinking of a less incisive control of the Church in the 5th century than one can commonly believe.
In the rotunda of sarcophagi is surprising the concentration of monumental burials, so that one can think members of the Church commissioned the works here too. What clues are the that could support the hypothesis that the clients belonged to a religious community? To tell the truth, they are scant and among them the epigraph of the blessed virgins Fotina and Filomena deserves to be recorded: the former lived 80 years, the latter 84 (IG XIV, 187; FERRUA 1989, 180). The sole dated inscription found within this chamber, which is related to the name of Eucarpio, bearing the notification of both the consuls between 339 and 360 (FERRUA 1983, 3), cannot be used for dating; the discovery data, insisting on the fact that the gravestone was found overturned in a forma tomb (AGNELLO 1960, 30-31), suggest that the gravestone was reused.

>>>>

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Towards the Publication of ICI Siracusa: General Data and Previews.}
\author[unica]{Mariarita Sgarlata\corref{first}}
\address[unica]{Department of Humanities - University of Catania} 
\cortext[first]{Corresponding author. Email: m.sgarlata@unict.it}
\begin{document}
\maketitle
\begin{abstract}
This paper refers to the general track ``Epigraphic edition on paper and on line''. Some researchers who has been working on the editing of the \emph{Inscriptiones Christianae Italiae}, published from the University of Bari, contributed also to the EDR project, that collect on line the epigraphic documents of Roman Christian Period. I propose a preview of the work currently in progress, with a specific reference to the inscriptions that provide us the chronological and topographic data to study the cemetery as well as formularies linked to the society structure and to the identity-making characteristics.
\end{abstract}
\keywords{Sicily, Syracuse, Epigraphy, Topography, Society, Onomastic and Identity}

@@

%\begin{figure}[hbp]
%\centering
 %\includegraphics[width=\columnwidth]{MISSING IMAGE!}
%\caption{The main gallery (so called \emph{decumanus}) of Catacombs of San Giovanni}
%\label{fig:1}
%\end{figure}

\noindent We are far from a complete and thorough research about palaeochristian epigraphs in Sicily. This catalogue represents the first step of a systematic study focused on the inscriptions discovered in the christian cemeteries of Siracusa.

@@

The epigraphic research related to San Giovanni catacomb can count on certain informations about the discovery place thanks to Paolo Orsi archeological campaigns. However, looking to previous publications, sometimes it’s necessary a review of the presentations made by Mommsen and Kaibel. In this perspective the main sources of the CIL and IG authors were verified so we had a deeper vision of the relationship between Mommsen and Kaibel and the main sources used, sometimes reevaluating the contribution of this sources. The renovated interest about funerary epigraphy generated new studies about new criteria of dating. The inscriptions dated by the indications of the ``consular couple'' represent an exceptions in the issues about chronological datings \citep[227]{FERRUA1947} \citep[nn. 191-208]{FERRUA1989}: two were discovered during the excavation conducted by Paolo Orsi (1985) and to them is associated the latin epigraph ``Sporus'' (356) .

The bigger one is entitled to San Giovanni and gave back almost 800 inscriptions, now located for the most part in the Archeological and Regional Museum ‘’Paolo Orsi’’ of Siracusa.; alongside will be placed the 40 headstones recovered in the adjacent hypogea of Villa Landolina, that report compatible datings and formularies.

\section{HISTORICAL ASPECTS OF THE CATACOMB OF S. GIOVANNI AT SYRACUSE}

\noindent The systematic studies of collective cemeteries in Syracuse started many years ago and, beyond the archaeological surveys, the late antique funerary settlements of Syracuse proposed multiple research cues and paths, as the historical- religious, economic and social nature.
Suburban cemeteries, fanned out from the area of Fusco, in the quarter of Neapolis, to the Santa Lucia area, in the southern part of Acradina, this indicates unequivocally what the perimeter of the city must have already been in the early and mid Roman Empire. The History of the area, which was going to hold the catacombs (San Giovanni, Vigna Cassia and Santa Lucia), spanned the centuries between the classical Greek and late antique ages, gradually giving evidence of quarries (Latomie), water supply systems to the city, characterized by cisterns and aqueducts \citep[682]{COLLINBOUFFIER1987}), handcraft workshops from the beginning of the 4th/3rd century BCE and burials datable to the early and mid Roman Empire. The analysis of the funerary system certifies one hand the dependence on the Roman model, and other the debt in respect of local traditions.

Several interest will be given to structural aspect of the catacomb of S. Giovanni, practice of funeral rituals, ethnic and cultural fruition’s characters, transformation in the use, transformation in the way of using spaces for graves, to complete a general point of view about the phenomena of continuity and innovation as to previous sepulchral arrangements and, in the analysed periods, the facies belonging to the different settling, variegated in the committees’ ideological and religious themes, in choosing monumental types (like the rotundae) and decorations, in self-representative aspects, in burial uses. In this perspective we will give particular importance to the study of funerary epigraphy aimed at the writing of ICI Siracusa and the overall interpretation of the monument.
Between 1893 and 1909 the archaeologist Paolo Orsi carried out a series of campaigns in the catacomb of San Giovanni. Detailed reports of those campaigns are recorded at various times in Notizie degli Scavi, which constitute an indefeasible starting point. The first incisive studies on Christian subterranean Sicily pertain to Joseph Führer, who dedicates many pages to the catacomb of San Giovanni, whose study is also epigraphic, is based on the previous literature, especially on Paolo Orsi’s discoveries \citep[13-39]{FUHRER1897}; \citep[22-26]{FUHRERSCHULTZE1907}; see also \citep[276, n. 2]{ORSI1893} 
%and \citep[60-61, fig. 5]{SCHULTZE1882} MISSING IN BIBLIOGRAPHY
. In the study of Syracuse cemeteries Orsi \citep[189]{ORSI1900} was the first to see the mixed nature of the burials and their materials (mostly gravestones), openly attesting a kind of pagan-Christian and orthodox-heterodox symbioses %\citep[242-243]{AGNELLO1957a} MISSING IN BIBLIOGRAPHY
. Both Antonio Ferrua and Santi Luigi Agnello within the space of a few years tried to solve the problems relating to the sarcophagus of Adelfia and the more general problem, strictly correlated with the ones just mentioned, of the cemetery genesis and development \citep{FERRUA1952} and \citep{AGNELLO1956}. It is necessary to record that numerous hypotheses on the genesis and development of the various sectors of the cemetery are ascribable to the account given by inscriptions. Syracuse, thanks to its prolific underground cemeteries, has the larger Christian epigraphic heritage after Rome, kept in the storage of the Soprintendenza di Siracusa.

\section{TOPOGRAPHIC AND EPIGRAFIC DATA TO STUDY THE CEMETERY}

\noindent Walking across the main gallery, one can retrace the stages of Paolo Orsi’s interventions, being at the entrance of the Catacomb second northern gallery, before the so-called ``tomb of the Saint''. This arcosolium is considered a privileged burial on the basis of the following reasons: most of all because of its position, but also because it is a mensatype burial closed by a sole slab \citep[292-294]{ORSI1893}. The signs of an ancient rite are easily traceable on the slab, a rite that preceded the coming of Christianity and persisted for centuries up to the present day: the rite of refrigerium, which literally means refreshment, cooling \citep{GIUNTELLA1985}. In the Christian ceremony the purpose of a funeral banquet is to benefit the soul of the departed on the anniversary of death, a painful event celebrated as dies natalis of the soul to eternal life.

Who was buried in this sepulchre? The question is destined to remain unanswered and only an inscription found nearby could be a clue in this sense. The text says that the owners purchased the sepulchre close to the one of the bishop Cheperion’s \citep[507-508]{ORSI1895}; \citep[55-58]{RIZZONE2011}, of whom the scanty written sources never make mention. In any case this constructed sarcophagus represent one of the most eminent burials of the catacomb. In the same gallery one cannot ignore the finding of an inscription in many fragments, which is a singular phenomenon of religious contamination.

This inscription, with a Christological monogram at the top, records Nassiana ``Christian, who competed with Penelope in moral virtue''. This inscription can only be compared with three examples in the Roman sepulchral carmina \citep{CLE}. The circular support of this inscription found close to the ``tomb of the Saint'' has been regarded as a mensa (table) for refrigerium rite \citep[47]{GIUNTELLA1985}, whose circular form would derive in any case from an evident reuse of a marble disk of classical craftsmanship, with a laurel wreath and berries sculpted on one side \citep[n. 234]{ORSI1895}. A wall inscription painted on the extrados of an arcosolium in the third northern gallery of the catacomb could be connected with Nassiana’s text. Two lines of the inscription say that ``Sossa outdid (other women) in conjugal love; as for handiworks, without being taught by anyone, Athena herself had taught her how to do marvellous things'' \citep[n. 3]{FERRUA1940}. Both inscriptions unequivocally make use of female figures from the classical world, interpreted as a model also for Christian women either for a deep-rooted usage or making up for the absence of assimilable figures in the Church. This allows one to notice in a 4th century large community cemetery, such as San Giovanni, some phenomena of religious contamination in a period in which Christian epigraphic praxis consolidated itself by then \citep[484]{SGARLATA1999}.

The rotunda derives its name from the deceased Antiochia recorded in the sarcophagus, set inside the ring of tombs, made of blocks and bearing an engraved and rubricated inscription. Whom did this private space belong to? To answer the question, one can rely on a suggestion provided once more by epigraphic testimonies. The gravestones found – on which they put as a rule the deceased’s name, lifespan, date of death and deposition – according to the first excavators’ reports \citep[23-45]{CARINI1873}, attest that the rotunda of Antiochia could have housed women only and this would give credence to the idea that the mausoleum had been used for a female monastic community. The hypothesis needs to be verified, but it is very seductive.

It is evident that the catacomb of San Giovanni was originally a community cemetery, planned for only one type of burial: the arcosolium with multiple depositions, which does not require great care for decoration, only transennae. In the topographical and architectural development of the catacomb it appears clear that creating the rotundas breaks up the common burials series, destined to a socially homogeneous Christian community. These changes to the original plan – of the creation of subterranean mausolea both to the north and south of the cemetery – spring from the necessity to create appropriate spaces for the members of the Chruch and above all of the Empire, bringing into question the initial egalitarian choice of the arcosolium burials \citep[767]{GRIESHEIMER1989}. In the terminal part of the pious Giovanni’s gallery a monumental sarcophagus, once more hewn out of the rock, can be perceived. Close to the closing wall of the same gallery an inscription was found: it bears the notification of both the consuls of the year 349 %\citep[756, 8]{ORSI1909} MISSING IN BIBLIOGRAPHY
; \citep[89]{AGNELLO1953}. As a rule this date marks the end of the dig works in this sector of the catacomb \citep[75-76]{FERRUA1952}. A arcosolium extrados, in the eastern region, shows traces of a palinsesto, an overlap of paintings and epigraphs than obstruct the identification of the first person buried in this sepulchre, even if recently Vittorio Rizzone proposed the interpretation Philadelpheia.\footnote{\citet[45-48]{RIZZONE2012}.} In one of the fossa tombs at the terminal stretch of the main gallery the inscription of Euterpe (IG XIV, 112) has been found, reused to cover the bottom of the tomb, which traditionally marks the end of the dig works of the catacomb \citep[75-76]{FERRUA1952}; \citep[79]{AGNELLO1958}; \citep[781]{GRIESHEIMER1989}. Beyond the deceased’s biometric data, recorded as ``companion of the Muses'', the epigraph in the last three lines mentions the consuls’ iteration, which allow us to date back to the consulate of the Emperor Constantius, consul for the tenth time, and Julian- Caesar, consul for the third time. So Euterpe died on November 27th 360 A.D. at the age of 22 years and 3 months \citep[524-526]{GUARDUCCI1978}.

In the southern region,, more than in other ones, that you can note the transformations that have profoundly undermined its community spirit, which had originally inspired the creation of the catacomb. In the first rotunda of the southern region, is a private space, which was given the name of Marina due to an inscription scratched upon the extrados of the arcosolium,on the right of the entrance to the short gallery of the bishop Siracosio. The arcosolium seems to be enframed by a painted prothyrum, as the still visible column and capital attest, confirming the generalized use of architectural elements in this catacomb, already seen in the rotunda of Antiochia. According to the interpretation of the text, Marina could have been the wife of the patriciuset magister militum Sabinianus, sent by the Emperor Honorius to Spain at the time of barbarian invasions presumably between 409 and 423 \citep[21-22, 40]{FERRUA1989}.

This date would agree with the numerous testimonies given by the inscriptions that have been dated thanks to the notification of the two consuls in office in the year of their death.These testimonies are ascribable to the years of the Emperors Arcadius and Honorius in the first quarter of the 5th century and seem to suggest a link between a few burials of the southern region and the aristocrats’ diaspora from Rome after Alaricus’ advance in 410, who took refuge in Sicily and Africa as they did in other provinces of the Empire \citep[175]{SIRAGO1989}; %\citep[157-158]{VERA1988} MISSING IN BIBLIOGRAPHY
. It is necessary to have a look at the slab of the presumed arcosolium of the bishop Siracosio, recorded in one inscription found in an adjacent fossa tomb, which says that the deceased intentionally purchased the sepulchre close to the one of the bishop just mentioned (IG XIV, 123: \begin{otherlanguage*}{greek}Ἐνθάδε κῖτε Πολυχ/ρόνιος καὶ Σεραπία. / Ἠγόρασεν τῷ τότ/ε καιρῷ / Πολυχρονίου /αἱ Σεραπία ἐπὶ τῷ κυρί/ῳ μου ἐπισκόπῳ Συρα/κοσίῳ\end{otherlanguage*}). It is just a hypothesis seeing in the arcosolium with the engraved slab, still in situ, a noble burial for a member of the ecclesiastical hierarchy, for whom, in the absence of other data, according to a letter of pope Gelasius I, an episcopate between 492 and 496 has been proposed \citep[223]{NARCISO1952} \citep{CARLETTI2008}. One can clearly distinguish a Christogram with the apocalyptic letters alpha and omega and two ships in the shape of fish, regarded as making reference to the sacrament of the Eucharist, this is also suggested by the diskettes next to the fishes’ mouths, assimilable to loaves of bread \citep{SGARLATA2013}.

After reading the text of the inscription upon the sarcophagus lid (CIL X, 7123: Ic Adelfia c(larissima) f(emina) / posita conpar / Balericomitis), one can become aware that it refers only to the woman, not to both husband and wife: here lies Adelfia, clarissima femina, wife of the count Valerius. Who were therefore Valerius and Adelfia? The unsolved enigmas of the catacomb of San Giovanni deal with their names and most of all their identification. In reality it is possible to distinguish more than two phases of intervention, which have preceded the creation of the hole for the sarcophagus. At the beginning the internal space of the large niche was scenographically arranged in a terrace pattern with a forceful ascensional effect, not exempt from comparisons \citep[14, plate VII]{NESTORI1973}.The intact Latin inscription of Sporus, dated to 356, which sealed a forma in the main gallery among the ones that join the rotunda of Marina to the one of Adelfia \citep[24]{CAVALLARI 1872}; \citep[90]{AGNELLO1953}, would attest that the exploitation of the soil in the area gravitating to the rotunda of Adelfia was already underway after the first half of the 4th century, so confirming the evident anteriority of the six graves cut in the floor to the phase of monumentalization. Both the internal and the external organizations of the large niche (nicchione) would correspond to the first and second phases of intervention respectively, according to the current reconstruction. The problems related to the reconstruction of the third phase seem to be more complicated: the monumentalization phase started with the interment of the sarcophagus and concluded with the acquisition of an aspect comparable to the privileged burials of Roman crypts.

The monument seems to have a less wild internal dynamics of development; the new data (archaeological, historical, epigraphical) \citep[101-108]{SGARLATA1996} allow the scholars to widen the syncopated weave of temporal sequences, to which the analysis of the monument has been pinned by the constant reference to Valerius Proculus’ chronology. A new chronology and a different identification of the comes Valerius can be proposed, considering: 1) the evident reuse of the sarcophagus; 2) the topographical development of the catacomb in the area where the sarcophagus has been discovered; 3) the type of monumental intervention, which followed the Roman counterparts, datable to the second half of the 4th century \citep[132-134]{FIOCCHINICOLAI 1997}. If Valerius were given a different identity, one could postpone him from the age of Constantine to that of Augustine \citep[15-51]{SGARLATA1998}, on the basis of several accounts about the friendship between the saint from Hippo and a comes Valerius, whose physiognomy, is somewhat vague. This friendship, confirmed by the epistles (AUG., Ep. 200, 206, 207; Retr. II, 79 and 88) and the dedication, in 419, of the treatise De nuptiis et concupiscentia, is fed on the fight against Pelagianism, which in eastern Sicily had found a fertile soil. The presence of Pelagians in Sicily is attested for certain, according to Hilarius’ account under the pontificate of Innocent \citep[429-452]{PIETRI2000} and Honoreficentia’s letter, where a clarissima based at Syracuse is mentioned. On the island the spread of Pelagian movement looks like a direct consequence of the 410 sacking of Rome and the diaspora of the Roman nobility, of which Pelagius and Celestius were spiritual leaders; the short period they both spent in Sicily was not painless for Christian orthodoxy %\citep[436-437]{PIETRI2001} MISSING IN BIBLIOGRAPHY
 and particularly in Syracuse, as the cemeteries in the area overhanging the Greek theater, intended to serve the communities of the so-called heretics throughout the 5th century \citep{AGNELLO1990}.All the data gathered seems to lead to a different identification of Valerius – who, even if he was not Augustine’s correspondent, would be sought in the list of Valerii reported in the sources of the first quarter of the 5th century (PLRE II, 1143-44) – and a later chronology of the large niche monumental transformation compared to the one traditionally accepted.

@@

To answer the questions will take time, which will assure a greater credibility to what so far seems just a series of suggestions, most of all fed by the re-reading of epigraphic testimonies: the episodic character of the references to burials of bishops, presbyters, deacons and noting that the most significant percentage of evidences pertains to the members of the Church, buried in Syracuse away from their own countries and recorded in wall inscriptions written in Latin – Auxentius Hispanus episcopus and Superianus clerecus de Aquileia \citep[1 and 6]{FERRUA1940} – as a demonstration that the official language is used by foreigners, who were high clients (the remainder of inscriptions is in Greek), these and many other clues lead to thinking of a less incisive control of the Church in the 5th century than one can commonly believe.

In the rotunda of sarcophagi is surprising the concentration of monumental burials, so that one can think members of the Church commissioned the works here too. What clues are the that could support the hypothesis that the clients belonged to a religious community? To tell the truth, they are scant and among them the epigraph of the blessed virgins Fotina and Filomena deserves to be recorded: the former lived 80 years, the latter 84 (IG XIV, 187); \citep[180]{FERRUA 1989}. The sole dated inscription found within this chamber, which is related to the name of Eucarpio, bearing the notification of both the consuls between 339 and 360 (\citep[3]{FERRUA1983}), cannot be used for dating; the discovery data, insisting on the fact that the gravestone was found overturned in a forma tomb \citep[30-31]{AGNELLO1960}, suggest that the gravestone was reused.
